% Heuristic-Guided Prompt Engineering -- NeurIPS-style extended abstract (1 page).
%
% Self-contained: compiles with a standard LaTeX install (pdflatex paper.tex);
% it does NOT depend on neurips_2024.sty. This is a non-anonymous extended
% abstract; for a double-blind track, replace the author block with
% "Anonymous Author(s)" and use the venue's submission (line-numbered) style.
\documentclass[10pt]{article}
\usepackage[letterpaper,margin=1in]{geometry}
\usepackage{times}
\usepackage{microtype}
\usepackage{amsmath}
\usepackage[hidelinks]{hyperref}
\usepackage{titlesec}
\titlespacing*{\section}{0pt}{1.2ex}{0.6ex}
\titlespacing*{\paragraph}{0pt}{0.6ex}{1em}
\setlength{\parskip}{0.15em}
\pagestyle{empty}

\title{\vspace{-3.5em}\textbf{Prompts that Learn: Heuristic-Guided Prompt
Engineering\\ via Expectation Maximization in LLM Workflows}}
\author{Mark Schutera\\\texttt{mark.schutera@mailbox.org}}
\date{}

\begin{document}
\maketitle\vspace{-2.2em}

\begin{abstract}
\noindent
We ask whether a \emph{small} (3B-parameter) local LLM, used as the per-car
driving policy with \emph{all} knowledge in the prompt, can match crowd-sourced
deep RL on MIT DeepTraffic. On a frame-exact re-implementation of the benchmark
(11-car, official $500\!\times\!2000$ deterministic protocol; score $=$ median
fleet mph), a prompt-only \texttt{qwen2.5:3b} policy reaches a 24-run median of
\textbf{76.5\,mph} (point estimate; expected median ${\sim}76.4$, $P({>}76.3)
{=}0.85$), with a mean (76.13) matching the best hand-tuned heuristic's 500-run
mean (76.12) and exceeding the top real DQN submission (74.13) by ${\sim}2.4$\,mph.
We then formalize the systematic \emph{manual} tuning that achieved this as
\textbf{prompt-EM}: a regression/EM loop that co-optimizes the \emph{weighting}
and the \emph{ordering} of a small set of human-written driving heuristics, which
an LLM synthesizes into one prompt, scored each round by the simulator's median
speed as a closed-loop feedback signal. Run at scale over 16 heuristics, prompt-EM recovers a strong
prompt automatically (proxy median 75.9, well above random search), the residual
gap to hand-tuning being prompt \emph{wording}. We release the suite with full
provenance and per-iteration logging so that heuristic importance and
\emph{sequence} importance can be analyzed.
\end{abstract}
\vspace{-0.6em}

\section{Setup}
DeepTraffic scores the median, over many deterministic episodes, of the average
speed of a fleet of identical agents weaving through slower traffic. Our engine
reproduces the original JavaScript bit-for-bit (18/18 fidelity tests pass; max
per-frame deviation $0.0$, single- and multi-agent). Reference points on this
engine: do-nothing floor ${\sim}60$; best real trained-DQN submission
\textbf{74.13}; archived leaderboard \#1 \textbf{75.66}; published all-time
ceiling \textbf{76.3}. The repo's prior-best heuristic sits \emph{below} this
ceiling (expected median 76.17); a new hand-coded convention sweep gives an
expected median of \textbf{76.40} ($5$\,SE above 76.3 over ten independent
500-run blocks), the first policy here to clear it. A second, more aggressive
sweep finds \emph{no} configuration above ${\sim}76.5$, so $76.3$ sits near the
engine's practical optimum for \emph{any} policy.

\section{A small LLM at the ceiling}
The policy receives a \emph{lossy semantic rendering} of the same observation the
DQN sees---the nearest car per lane within the ego's lane and its two neighbours,
as ``open / fast-far / slow-close'' plus an explicit rear status---and
a short prose driving convention; it reasons one sentence and emits one of five
actions. At temperature $0$ it is a deterministic function of the prompt, so an
exact-text cache is decision-identical to querying every step (verified: pure
uncached runs reproduce cached scores exactly). The tuned prompt reaches a
\textbf{24-run median of 76.50} (mean 76.13, matching the heuristic's 500-run mean
76.12); a bootstrap places $P(\mathrm{median}>76.3)\approx0.85$. Three findings
drove this from a 72.3 naive baseline: (i) a 3B model follows a \emph{prose}
convention but fails numbered numeric rule-lists; (ii) agreement with the
heuristic oracle does \emph{not} track score---\emph{blocked-state} decision
quality does; (iii) small wording and \emph{ordering} changes swing several mph,
because the model anchors on whichever rule is stated first/most forcefully.

\section{Prompt-EM}
We systematize this. Let $h_{1{:}K}$ be atomic heuristics, each a
one-sentence rule. A genome carries two simplex vectors: emphasis $\alpha$ (which
rules, how strongly) and priority $\rho$ (their order). Each iteration: (1) an LLM
\textbf{synthesizes} one system prompt that states the heuristics in
$\rho$-order, stressing high-$\alpha$ ones (a deterministic template is the
fallback); (2) \textbf{evaluate} over 10 runs $\rightarrow$ median; (3)
\textbf{E-step}: ridge-regress the objective on past $\alpha$ (and on $\rho$) to
estimate each heuristic's marginal contribution and its order-contribution;
(4) \textbf{M-step}: $\alpha\!\leftarrow\!(1{-}\eta)\,\alpha_{\text{best}}+\eta\,
r(\alpha)$, moving toward the positive-contribution \emph{mixture} (not the single
best rule), with learning rate $\eta$ and annealed exploration. A Latin-hypercube
cold start makes the otherwise-collinear regression identifiable (synthetic tests
recover the true helpful/harmful rules in ${\sim}12$ iterations). Single
heuristics map to prose \emph{1-to-1} (no LLM); only the weighted \emph{blend}
uses LLM synthesis. The state renderer is held fixed---we optimize the rules, not
the encoding.

\paragraph{Prompt-EM in practice.}
A 3B model is a high-variance \emph{synthesizer}: free-form LLM synthesis of the
blended prompt collapses the policy into a passive ``maintain'' mode (40--74 mph)
because it fails to reliably state the dominant ``accelerate-by-default'' rule.
Baking the proven wording into the heuristics and using the deterministic
template removes this variance (stable 74.6). The EM cold start then doubles as
an \emph{automatic ablation}: zeroing the accelerate rule's weight collapses the
fleet to 45 mph, cleanly marking it the dominant heuristic. Over $K{=}16$
heuristics the exploit phase reaches a 1000-frame-proxy median of \textbf{75.9}
(mean 75.0 vs.\ 72.9 for the random Latin-hypercube cold-start samples): at this
budget the loop's gain over random search is \emph{reliability}---it concentrates
in the good region---rather than a higher peak.

\paragraph{Positioning.}
Prompt-EM is black-box prompt optimization (cf.\ APE~\cite{ape}, OPRO~\cite{opro},
EvoPrompt~\cite{evoprompt}, Promptbreeder~\cite{promptbreeder},
APO/ProTeGi~\cite{apo}, TextGrad~\cite{textgrad}, DSPy~\cite{dspy}) applied to an
LLM-as-driving-policy (cf.\ DiLu~\cite{dilu}, GPT-Driver~\cite{gptdriver},
LanguageMPC~\cite{languagempc}). Its distinguishing combination is an
\emph{interpretable} search space---a weight-and-order simplex over human
heuristics, not free-form or soft-prompt text---updated by a \emph{regression/EM}
credit assignment that reports \emph{which} rules and \emph{which positions} the
loop's feedback favors; instruction-order credit is absent from prior black-box
prompt optimizers, and the objective is closed-loop multi-agent control.

\section{Limitations}
The margin over 76.3 is necessarily thin (the engine ceiling is ${\sim}76.5$) and
our 24-run significance is suggestive (85\%), not yet 95\%. Credit assignment over
$K$ heuristics from few in-loop evaluations is sample-limited, and emphasis and
order confound under joint search. The policy and the DQN share an observation
\emph{source} but not identical input features: the renderer is fixed and
information-non-increasing---no privileged data reaches the policy---yet it
exposes only a hand-thresholded slice of the 315-cell grid (nearest car per lane,
ego $\pm 1$ lanes), so part of the policy's competence sits in this feature map
rather than in the model. The automated loop reaches the mid-75s without
any manual wordsmithing but does not yet match the hand-tuned 76.5: the residual
gap is prompt \emph{wording}, which neither the weight nor the order channel
reaches. Closing it---a staged synthesizer \emph{freedom} that explores wording
around a frozen best structure---and isolating sequence importance with a
dedicated order-only run are immediate future work; both are implemented as ready
hooks. All runs ship with a full reproducibility manifest and per-iteration logs
(weights, order, action distributions, contributions).

{\small
\begin{thebibliography}{99}
\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}
\bibitem{ape} Zhou et al. Large Language Models Are Human-Level Prompt Engineers (APE). arXiv:2211.01910, 2022.
\bibitem{opro} Yang et al. Large Language Models as Optimizers (OPRO). arXiv:2309.03409, 2023.
\bibitem{evoprompt} Guo et al. EvoPrompt. arXiv:2309.08532, 2023.
\bibitem{promptbreeder} Fernando et al. Promptbreeder. arXiv:2309.16797, 2023.
\bibitem{apo} Pryzant et al. Automatic Prompt Optimization (APO/ProTeGi). arXiv:2305.03495, 2023.
\bibitem{textgrad} Yuksekgonul et al. TextGrad. arXiv:2406.07496, 2024.
\bibitem{dspy} Khattab et al. DSPy. arXiv:2310.03714, 2023.
\bibitem{dilu} Wen et al. DiLu. arXiv:2309.16292, 2023.
\bibitem{gptdriver} Mao et al. GPT-Driver. arXiv:2310.01415, 2023.
\bibitem{languagempc} Sha et al. LanguageMPC. arXiv:2310.03026, 2023.
\end{thebibliography}
}

\end{document}
